\chapter{Datatype-Generic Programming}
\label{chap-dat-gen-program}

The implementation in \Cref{chap-spec-impl} is an efficient implementation for incrementally computing the summation over a \texttt{Tree} datatype. However, when we want to implement this functionality for a different datatype, a lot of code needs to be copied while the process remains the same. This results in poor maintainability, is error-prone and is in general boring work.

An example of reducing manual implementations for datatypes is the \textit{deriving} mechanism in Haskell. The built-in classes of Haskell, such as \texttt{Show, Ord, Read}, can be derived for a large class of datatypes. However, \textit{deriving} is not supported for recursive datatypes. Therefore, we use \textit{Datatype-Generic Programming}\cite*{gibbons2006datatype} to define functionality for a large class of datatypes.

In this chapter, we introduce Datatype-Generic Programming, also known as \textit{generic programming} in Haskell, as a technique that uses the structure of a datatype to define functions for a large class of datatypes. This prevents the need to write the previously defined functionality for every datatype.

\section{Introduction}
There are multiple generic programming libraries, however to demonstrate the workings of generic programming we will be using a single library as inspiration, named \texttt{regular}\cite*{regular2022}. Here the generic representation of a datatype is called a \textit{pattern functor}. A pattern functor is a stripped-down version of a datatype, by only containing the constructor but not the recursive structure. The recursive structure is done explicitly by using a fixed-point operator. 

First, the pattern functor defined in \texttt{regular} are 5 primitive type constructors and 2 meta information constructors. The primitive type constructors describe the datatypes. The meta information constructors only contain information (e.g., constructor name) but not any structural information.

\pagebreak
\begin{minted}{haskell}
data U r         = U                 -- Empty constructor
data I r         = I r               -- Recursive call
data K a r       = K a               -- Constant
data (f :+: g) r = L (f r) | R (g r) -- Sums (Choice)
data (f :*: g) r = (f r) :*: (g r)   -- Products (Combine)
\end{minted}

The conversion from datatypes into pattern functors is done by the \texttt{Regular} type class. The \texttt{Regular} type class has two functions. The \texttt{from} function converts the datatype into a pattern functor and the \texttt{to} function converts the pattern functor back into a datatype. In \texttt{regular}, the pattern functor is represented by a type family. Then using the \texttt{Regular} conversion to a pattern functor, we can write the \texttt{Tree} datatype from \Cref{chap-spec-impl} as:

\begin{minted}{haskell}
type family PF a :: * -> *

class Regular a where
  from :: a -> PF a a
  to   :: PF a a -> a

type instance PF (Tree a) = K a                -- Leaf
                         :+: (I :*: K a :*: I) -- Node

class Regular (Tree a) where
  from (Leaf x)     = L (K x)
  from (Node l x r) = R (I l :*: K x :*: I r)                 

  to (L (K x))                 = Leaf x
  to (R (I l :*: K x :*: I r)) = Node l x r
\end{minted}

To demonstrate the workings of generic programming, we are going to implement a simple generic function which determines the length of an arbitrary datatype. First, we define the length function within a type class. The type class is used, to define how to compute the length for every primitive type constructor \texttt{f}. 

\begin{minted}{haskell}
class GLength f where
  glength :: (a -> Int) -> f a -> Int
\end{minted}

Writing instances for the empty constructor \texttt{U} and the constants \texttt{K} is simple because both primitive type constructors return zero. The \texttt{U} returns zero, because it does not contain any children. The \texttt{K} returns zero, because we do not count constants for the length. 

\begin{minted}{haskell}
instance GLength U where
  glength _ _ = 0

instance GLength (K a) where
  glength _ _ = 0
\end{minted}

The instances for sums and products are quite similar. The sums recurses into the specified choice. The product recurses in both constructors and combines them.

\begin{minted}{haskell}
instance (GLength f, GLength g) => GLength (f :+: g) where
  glength f (L x) = glength f x
  glength f (R x) = glength f x

instance (GLength f, GLength g) => GLength (f :*: g) where
  glength f (x :*: y) = glength f x + glength f y
\end{minted}

The instance for the recursive call \texttt{I} needs an additional argument. Because, we do not know the type of \texttt{x}, so, an additional function (\inlinehaskell{f :: a -> Int}) needs to be given which converts \texttt{x} into the length for that type.

\begin{minted}{haskell}
instance GLength I where
  glength f (I x) = f x
\end{minted}

Then using the \texttt{GLength} instances for all primitive type constructors, a function can be defined using the generic length function. By first, converting the datatype into a generic representation, then calling \texttt{glength} with the function \texttt{length} recursively, and for every recursive call increase the length by one. 

\begin{minted}{haskell}
length :: (Regular a, GLength (PF a)) => a -> Int
length = 1 + glength length (from x)
\end{minted}

\begin{minted}{haskell}
> length [1, 2, 3]
    3
> length (Node (Leaf 1) 2 (Leaf 3))
    3
> length {"1": 1, "2": 2, "3": 3}
    3
\end{minted}

\input{sections/generic_programming/explicit_recursion.tex}

\input{sections/generic_programming/sums_of_products.tex}

\input{sections/generic_programming/mult_rec_datatypes.tex}